---
editor_options: 
  chunk_output_type: console
excerpt: Introduction to Remote Sensing
---
### Load packages
```{r, title = "{{ site.handouts[0] }}"}
library(raster) # raster functionalities
library(rgdal) # GDAL/OGR binding for R with functionalities
library(sf) # spatial objects classes
library(colorRamps) # contains matlab.like color palette
library(psych) # pca/eigenvector decomposition functionalities
library(plotrix) # various graphic functions e.g. draw.circle

```

REB Notes: 
over plotting of the evergreen forest, urban and water
line 491: where does histogram() function come from???  replaced with raster::hist() 
overplotting of legend in time series pca


### Define functions used in this lesson
```{r}
create_dir_fun <- function(outDir, out_suffix=NULL){
  # if out_suffix is not null then append out_suffix string
  if(!is.null(out_suffix)){
    out_name <- paste("output_", out_suffix,sep="")
    outDir <- file.path(outDir, out_name)
  }
  # create if does not exists
  if(!file.exists(outDir)){
    dir.create(outDir)
  }
  return(outDir)
}
```

### Set up arguments and parameters
```{r}
#in_dir_reflectance <- "data/reflectance_RITA"
in_dir_var <- "../data"
out_dir <- "."
```

Region coordinate reference system taken from:
http://spatialreference.org/ref/epsg/nad83-texas-state-mapping-system/proj4/
```{r}
CRS_reg <- "+proj=lcc +lat_1=27.41666666666667 +lat_2=34.91666666666666 +lat_0=31.16666666666667 +lon_0=-100 +x_0=1000000 +y_0=1000000 +ellps=GRS80 +datum=NAD83 +units=m +no_defs" 

file_format <- ".tif" # Output format for raster images that are written out.

NA_flag_val <- -9999

out_suffix <-"exercise5_03312018" # Output suffix for the files and output folder # PARAM 8

create_out_dir_param <- TRUE 
```

### Define input data file names
```{r}
infile_reg_outline <- "new_strata_rita_10282017" # Region outline and FEMA zones

infile_modis_bands_information <- "df_modis_band_info.txt" # MOD09 bands information.

nlcd_2006_filename <- "nlcd_2006_RITA.tif" # NLCD2006 Land cover data aggregated at ~ 1km.

infile_name_nlcd_legend <- "nlcd_legend.txt" # Legend information for 2006 NLCD.

# MOD09 surface reflectance product on 2005-09-22 or day of year 2005265
infile_reflectance_date1 <- "mosaiced_MOD09A1_A2005265__006_reflectance_masked_RITA_reg_1km.tif"

# MOD09 surface reflectance product on 2005-09-30 or day of year 2005273
infile_reflectance_date2 <- "mosaiced_MOD09A1_A2005273__006_reflectance_masked_RITA_reg_1km.tif"
```

Create an ouput directory:
```{r}
if(is.null(out_dir)){
  out_dir <- dirname(in_dir) # output will be created in the input dir
}

out_suffix_s <- out_suffix # can modify name of output suffix
if(create_out_dir_param==TRUE){
  out_dir <- create_dir_fun(out_dir,out_suffix_s)
  setwd(out_dir)
}else{
  setwd(out_dir) #use previoulsy defined directory
}
```

## Part I: Display and explore the data

Read in MOD09 reflectance images before and after Hurrican Rita and do some basic plotting.
```{r}
r_before <- brick(file.path(in_dir_var, infile_reflectance_date1)) # Before RITA, Sept. 22, 2005.
r_after <- brick(file.path(in_dir_var, infile_reflectance_date2)) # After RITA, Sept 30, 2005.

plot(r_before) # Note that this is a multibands image.

reg_sf <- st_read(file.path(in_dir_var, infile_reg_outline))
reg_sf <- st_transform(reg_sf, crs=CRS_reg)
reg_sp <-as(reg_sf, "Spatial") # convert to sp object before rasterization

r_ref <- rasterize(reg_sp,
                   r_before,
                   field="OBJECTID_1",
                   fun="first")
plot(r_ref) # zone 2 is flooded and zone 1 is not flooded
```

Let's examine a location within FEMA flooded zone and outside: use centroids
```{r}
centroids_sf <- st_centroid(reg_sf)

df_before <- extract(r_before,centroids_sf)
df_after <- extract(r_after,centroids_sf)
```

Plot values of bands before and after for flooded region:
```{r}
plot(df_before[2,],type="l")
lines(df_after[2,],col="red")
```
Read band information since it is more informative!!
```{r}
df_modis_band_info <- read.table(file.path(in_dir_var,infile_modis_bands_information),
                                 sep=",",
                                 stringsAsFactors = F)
print(df_modis_band_info)

names(r_before) <- df_modis_band_info$band_name
names(r_after) <- df_modis_band_info$band_name
```

Order band in terms of wavelenth:
```{r}
df_modis_band_info <- df_modis_band_info[order(df_modis_band_info$start_wlength),]
# SWIR1 (1230–1250 nm), SWIR2 (1628–1652 nm) and SWIR3 (2105–2155 nm)
band_refl_order <- df_modis_band_info$band_number

plot(df_before[2,band_refl_order],type="l",
     xlab="Bands",
     ylab="Reflectance",
     main="Reflectance profile for centroid of flooded area")

lines(df_after[2,band_refl_order],col="red")

names_vals <- c("Before Rita","After Rita")

legend("topleft",legend=names_vals,
       pt.cex=0.7,cex=0.7,col=c("black","red"),
       lty=1, #add circle symbol to line
       bty="n")
```


## Part II: Examine spectral class signatures for land cover classes (from NLCD)

Read in NLCD 2006; note that it was aggregated at 1km.
```{r}
nlcd2006_reg <- raster(file.path(in_dir_var, nlcd_2006_filename))
```

Extract zonal averages by NLCD classes using zonal raster
```{r}
avg_reflectance_nlcd <- as.data.frame(zonal(r_before,nlcd2006_reg,fun="mean"))
```

Read in NLCD legend to add to the average information
```{r}
lc_legend_df <- read.table(file.path(in_dir_var,infile_name_nlcd_legend),
                           stringsAsFactors = F,
                           sep=",")

print(avg_reflectance_nlcd)

names(lc_legend_df)
```

Add relevant categories and plot
```{r}
lc_legend_df_subset <- subset(lc_legend_df,select=c("ID","NLCD.2006.Land.Cover.Class"))
names(lc_legend_df_subset) <- c("ID","cat_name")

avg_reflectance_nlcd <- merge(avg_reflectance_nlcd,lc_legend_df_subset,by.x="zone",by.y="ID",all.y=F)
head(avg_reflectance_nlcd)

names(avg_reflectance_nlcd)
col_ordering <- band_refl_order + 1

plot(as.numeric(avg_reflectance_nlcd[9,col_ordering]),
     type="l",
     col="green",
     xlab="bands",
     ylab="reflectance") #42 evergreen forest
lines(as.numeric(avg_reflectance_nlcd[6,col_ordering]),
      type="l",
      col="darkred") #22 developed,High intensity
```


## Part II: Explore data in feature space
```{r}
plot(r_before$Red, r_before$NIR)
plot(r_before$Green, r_before$Red)
plot(r_before$SWIR1, r_before$NIR)
plot(r_before$Red, r_before$SWIR1)

df_r_before_nlcd <- as.data.frame(stack(r_before,nlcd2006_reg))
head(df_r_before_nlcd)
```

### Feature space NIR1 and Red

Evergreen Forest:
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==42)

plot(df_subset$Green,
     df_subset$Red,
     col="green",
     cex=0.15,
     xlab="Green",
     ylab="Red",
     main="Green-Red feature space")
```

Urban: dense
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==22)

points(df_subset$Green,
       df_subset$Red,
       col="darkred",
       cex=0.15)
```

Water: 11 for NLCD
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==22)

points(df_subset$Green,
       df_subset$Red,
       col="blue",
       cex=0.15)

title("Feature space Green-Red")
names_vals <- c("Evergreen Forest","Developed High Intensity","Water")

legend("topright",legend=names_vals,
       pt.cex=0.7,cex=0.7,col=c("green","darkred","blue"),
       pch=1, # add circle symbol to line
       bty="n")
```

### Feature space Red and NIR

Evergreen Forest:
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==42)

plot(df_subset$Red,
     df_subset$NIR,
     col="green",
     cex=0.15,
     xlab="Red",
     ylab="NIR")
```

Developed, High Intensity
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==22)

points(df_subset$Red,
       df_subset$NIR,
       col="darkred",
       cex=0.15)
```

Water
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==11)

points(df_subset$Red,
       df_subset$NIR,
       col="blue",
       cex=0.15)

title("Feature space Red-NIR")
names_vals <- c("Evergreen Forest","Developed High Intensity","Water")

legend("topright",legend=names_vals,
       pt.cex=0.7,cex=0.7,col=c("green","darkred","blue"),
       pch=1, # add circle symbol to line
       bty="n")
```

### Feature space Blue and Red

Water:
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==11)

plot(df_subset$Blue,
       df_subset$Red,
       col="blue",
       cex=0.15,
       xlab="Blue",
       ylab="Red")
```

Evergreen Forest:
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==42)

points(df_subset$Blue,
       df_subset$Red,
       col="green",
       cex=0.15)
```

Urban dense:
```{r}
df_subset <- subset(df_r_before_nlcd,nlcd_2006_RITA==22)

points(df_subset$Blue,
       df_subset$Red,
       col="darkred",
       cex=0.15)

title("Feature space Blue-Red")
names_vals <- c("Evergreen Forest","Developed High Intensity","Water")

legend("topright",legend=names_vals,
       pt.cex=0.7,cex=0.7,col=c("green","darkred","blue"),
       pch=1, # add circle symbol to line
       bty="n")
```

### Generate Color composites

True color composite
```{r}
plotRGB(r_before,
        r=1,
        g=4,
        b=3,
        scale=0.6,
        stretch="hist")
```

False color composite:
```{r}
plotRGB(r_before,
        r=2,
        g=1,
        b=4,
        scale=0.6,
        stretch="hist")

plotRGB(r_after,
        r=2,
        g=1,
        b=4,
        scale=0.6,
        stretch="hist")
```
Note: the effect of flooding is particularly visible in the false color composites.
Let's examine different band combinations to enhance features in the image, including flooded areas.

## Part III: Band combination - indices and thresholding for flood mapping

NIR experiment with threshold to map flooding (water)
```{r}
plot(subset(r_before,"NIR"))
plot(subset(r_after,"NIR"))
```

Lower NIR often correlates to areas with high water fraction or inundated:
```{r}
r_rec_NIR_before <- subset(r_before,"NIR") < 0.2
r_rec_NIR_after <- subset(r_after,"NIR") < 0.2

plot(r_rec_NIR_before, main="Before RITA, NIR > 0.2")
plot(r_rec_NIR_after, main="After RITA, NIR > 0.2")
```
This is suggesting flooding!!!

Compare to actual flooding data:
```{r}
freq_fema_zones <- as.data.frame(freq(r_ref))
xtab_threshold <- crosstab(r_ref,r_rec_NIR_after,long=T)

## % overlap between the flooded area and values below 0.2 in NIR
(xtab_threshold[5,3]/freq_fema_zones[2,2])*100 
```
There is in agreement with FEMA flooded area in %.

### Generating indices based on raster algebra of original bands

Let's generate a series of indices, we list a few possibility from the literature.
1) NDVI = (NIR - Red)/(NIR+Red)
2) NDWI = (Green - NIR)/(Green + NIR)
3) MNDWI = Green - SWIR2 / Green + SWIR2
4) NDWI2 (LSWIB5) =  (NIR - SWIR1)/(NIR + SWIR1)
5) LSWI (LSWIB5) =  (NIR - SWIR2)/(NIR + SWIR2)

```{r}
names(r_before)
r_before_NDVI <- (r_before$NIR - r_before$Red) / (r_before$NIR + r_before$Red)
r_after_NDVI <- subset(r_after,"NIR") - subset(r_after,"Red")/(subset(r_after,"NIR") + subset(r_after,"Red"))

plot(r_before_NDVI,zlim=c(-1,1),col=matlab.like(255))
plot(r_after_NDVI,zlim=c(-1,1),col=matlab.like2(255))
```
Experiment with different threshold values: this is suggesting flooding!!!

Examine the histograms to look for potential threshold values.
```{r}
histogram(r_before_NDVI)  
histogram(r_after_NDVI) 

plot(r_before_NDVI < -0.5)
plot(r_after_NDVI < -0.5)
plot(r_before_NDVI < -0.1)
plot(r_after_NDVI < -0.1)
```

2) NDWI = (Green - NIR)/(Green + NIR)
3) MNDWI = Green - SWIR2 / Green + SWIR2

Modified NDWI known as MNDWI : Green - SWIR2/(Green + SWIR2)

According to Gao (1996), NDWI is a good indicator for vegetation liquid water content and 
is less sensitive to atmospheric scattering effects than NDVI. In some studies, MODIS band 6 
is used for the NDWI calculation, because it is sensitive to water types and contents 
(Li et al., 2011), while band 5 is sensitive to vegetation liquid water content (Gao, 1996).

```{r}
r_before_MNDWI <- (r_before$Green - r_before$SWIR2) / (r_before$Green + r_before$SWIR2)
r_after_MNDWI <- (r_after$Green - r_after$SWIR2) / (r_after$Green + r_after$SWIR2)

plot(r_before_MNDWI,zlim=c(-1,1),col=rev(matlab.like(255)))
plot(r_after_MNDWI,zlim=c(-1,1),col=rev(matlab.like(255)))

plot(r_before_MNDWI > 0.5)
plot(r_after_MNDWI > 0.5)
plot(r_before_MNDWI > 0.1)
plot(r_after_MNDWI > 0.1)
```
This is suggesting flooding!!!

```{r}
r_diff_MNDWI <- r_after_MNDWI - r_before_MNDWI

plot(r_diff_MNDWI)
plot(r_diff_MNDWI > 0.2)  
```
Increase in MNDWI indicates areas that may have become flooded.

Standardize image:
```{r}
mean_val <- cellStats(r_diff_MNDWI,"mean")
sd_val <- cellStats(r_diff_MNDWI,"sd")

r_diff_std <- (r_diff_MNDWI - mean_val)/ sd_val

plot(r_diff_std)
raster::hist(r_diff_std) # histogram(r_diff_std) 
plot(r_diff_std > 2) 
plot(r_diff_std > 1)  
```
Note the increase in MNDWI 2 standard deviations away


Generate a map of flooding with MNDWI and compare to FEMA map:
```{r}
r_after_flood <- r_diff_std > 1 

plot(r_after_flood)

# reclass in zero/1!!!

df <- data.frame(id=c(1,2), v=c(0,1))
r_ref_rec <- subs(r_ref, df)

plot(r_ref_rec)
xtab_tb <- crosstab(r_after_flood,r_ref_rec)
```

Examine overlap between flooded area and FEMA:
```{r}
plot(r_after_flood)

r_overlap <- r_ref_rec * r_after_flood

plot(r_overlap) # Flood map area of aggreement with FEMA
```


## Part IV: Principal Component Analysis and band transformation

Another avenue to combine original bands into indices is to use linear transformation.
We examine Principal Component Analysis and the Tassel Cap Transform.

Correlate long term mean to PC:
```{r}
cor_mat_layerstats <- layerStats(r_before, 'pearson', na.rm=T)
cor_matrix <- cor_mat_layerstats$`pearson correlation coefficient`
class(cor_matrix)
print(cor_matrix) # note size is 7x7

pca_mod <- principal(cor_matrix,nfactors=7,rotate="none")
class(pca_mod$loadings)
print(pca_mod$loadings)

plot(pca_mod$loadings[,1][band_refl_order],type="b",
     xlab="time steps",
     ylab="PC loadings",
     ylim=c(-1,1),
     col="blue")
lines(-1*(pca_mod$loadings[,2][band_refl_order]),type="b",col="red")
lines(pca_mod$loadings[,3][band_refl_order],type="b",col="black")
title("Loadings for the first three components using T-mode")
```

Make this a time series:
```{r}
loadings_df <- as.data.frame(pca_mod$loadings[,1:7])

title("Loadings for the first three components using T-mode")
names_vals <- c("pc1","pc2","pc3")
legend("topright",legend=names_vals,
       pt.cex=0.8,cex=1.1,col=c("blue","red","black"),
       lty=c(1,1), # set legend symbol as lines
       pch=1, #add circle symbol to line
       lwd=c(1,1),bty="n")
```

Add scree plot:
```{r}
plot(pca_mod$values,main="Scree plot: Variance explained",type="b")
```

Generate scores from eigenvectors:
Using predict function: this is recommended for raster.
```{r}
r_pca <- predict(r_before, pca_mod, index=1:7, filename="pc_scores.tif", overwrite=T) # fast

plot(r_pca,y=2, zlim=c(-2,2))
plot(r_pca,y=1, zlim=c(-2,2))
plot(r_pca,y=3, zlim=c(-2,2))

plot(subset(r_pca,1), subset(r_pca, 2))
plot(subset(r_pca,2), subset(r_pca, 3))
```


Generate a plot for PCA with loadings and compare to Tassel Cap:
```{r}
var_labels <- rownames(loadings_df)

plot(loadings_df[,1],loadings_df[,2],
     type="p",
     pch = 20,
     col ="blue",
     xlab=names(loadings_df)[1],
     ylab=names(loadings_df)[2],
     ylim=c(-1,1),
     xlim=c(-1,1),
     axes = FALSE,
     cex.lab = 1.2)
axis(1, at=seq(-1,1,0.2),cex=1.2)
axis(2, las=1,at=seq(-1,1,0.2),cex=1.2) # "1' for side=below, the axis is drawn on the right at location 0 and 1

box()    # This draws a box...

title(paste0("Loadings for component ", names(loadings_df)[1]," and " , names(loadings_df)[2] ))
draw.circle(0,0, c(1.0,1.0), nv=500) # ,border="purple",
text(loadings_df[,1], loadings_df[,2], var_labels, pos=1,cex=1)            
grid(2,2)
```

Plot feature space:
```{r}
df_raster_val <- as.data.frame(stack(r_after, r_pca, nlcd2006_reg))
head(df_raster_val)

# Water
plot(df_raster_val[df_raster_val$nlcd_2006_RITA==11,c("pc_scores.1")],
     df_raster_val[df_raster_val$nlcd_2006_RITA==11,c("pc_scores.2")],
     col="blue",cex=0.15,
     ylim=c(-2,2),
     xlim=c(-2,2))

# Urban: dense
points(df_raster_val[df_raster_val$nlcd_2006_RITA==22,c("pc_scores.1")],
       df_raster_val[df_raster_val$nlcd_2006_RITA==22,c("pc_scores.2")],
       col="brown",cex=0.15)

# Forest:
points(df_raster_val[df_raster_val$nlcd_2006_RITA==42,c("pc_scores.1")],
       df_raster_val[df_raster_val$nlcd_2006_RITA==42,c("pc_scores.2")],
       col="green",cex=0.15)

plot(r_pca$pc_scores.2 > 0.1) # water related?
plot(r_pca$pc_scores.2 < - 0.1) # vegetation related?
```
Note the negative values are related to Forest and positve to water on PC2


Examine relationship with Tasselcap transform adapted for MODIS:
```{r}
# 6) TCWI =  0.10839 * Red+ 0.0912 * NIR +0.5065 * Blue+ 0.404 * Green 
#            - 0.241 * SWIR1- 0.4658 * SWIR2-
#           0.5306 * SWIR3
# 7) TCBI = 0.3956 * Red + 0.4718 * NIR +0.3354 * Blue+ 0.3834 * Green
#           + 0.3946 * SWIR1 + 0.3434 * SWIR2+ 0.2964 * SWIR3

r_TCWI =  0.10839 * r_before$Red + 0.0912 * r_before$NIR +0.5065 * r_before$Blue+ 0.404 * r_before$Green 
- 0.241 * r_before$SWIR1- 0.4658 * r_before$SWIR2 - 0.5306 * r_before$SWIR3

r_TCBI = 0.3956 * r_before$Red + 0.4718 * r_before$NIR +0.3354 * r_before$Blue+ 0.3834 * r_before$Green
          + 0.3946 * r_before$SWIR1 + 0.3434 * r_before$SWIR2+ 0.2964 * r_before$SWIR3

# r_TCWI <- r_TCWI*10
raster::hist(r_TCWI) # histogram(r_TCWI)
plot(r_TCWI)
plot(r_TCWI, zlim=c(0,0.2))

plot(r_TCWI>0.1)

raster::hist(r_TCBI) # histogram(r_TCBI)
plot(r_TCBI)
plot(r_TCBI,zlim=c(0, 0.2))

r_stack <- stack(r_TCWI, r_TCBI, r_pca)
names(r_stack) <- c("TCWI","TCBI", names(r_pca))
cor_r_TC <- layerStats(r_stack, "pearson", na.rm=T)
cor_r_TC <- as.data.frame(cor_r_TC$`pearson correlation coefficient`)
print(cor_r_TC)
```

Plot on the loading space using TCBI and TCWI as supplementary variables:
```{r}
var_labels <- rownames(loadings_df)

plot(loadings_df[,1],loadings_df[,2],
     type="p",
     pch = 20,
     col ="blue",
     xlab=names(loadings_df)[1],
     ylab=names(loadings_df)[2],
     ylim=c(-1,1),
     xlim=c(-1,1),
     axes = FALSE,
     cex.lab = 1.2)

points(cor_r_TC$pc_scores.1[1], cor_r_TC$pc_scores.2[1], col="red")
points(cor_r_TC$pc_scores.1[2], cor_r_TC$pc_scores.2[2], col="green")

axis(1, at=seq(-1,1,0.2), cex=1.2)
axis(2, las=1,at=seq(-1,1,0.2), cex=1.2) # "1' for side=below, the axis is drawn on the right at location 0 and 1

box()    # This draws a box...

title(paste0("Loadings for component ", names(loadings_df)[1]," and " ,names(loadings_df)[2] ))
draw.circle(0,0,c(1.0,1.0),nv=500) # ,border="purple",
text(loadings_df[,1],loadings_df[,2],var_labels,pos=1,cex=1)            
text(cor_r_TC$pc_scores.1[1],cor_r_TC$pc_scores.2[1],"TCWI",pos=1,cex=1)
text(cor_r_TC$pc_scores.1[2],cor_r_TC$pc_scores.2[2],"TCBI",pos=1,cex=1)
grid(2,2)


```
