---
editor_options: 
  chunk_output_type: console
excerpt: Raster Classification using flood mapping analysis
---

```{r, include=FALSE, echo=FALSE, results='hide'}
library(raster) # raster functionalities
library(sf) # spatial objects classes
library(rpart) # regression and classification trees 
library(e1071) # support vector machine
library(rasterVis) # raster visualization operations
library(caret) # complex regression and classification models
```


### Set up arguments and parameters

```{r, title = '{{ site.handouts[0] }}'}
in_dir <- "data"

out_dir_suffix <- "rsc_lesson"
out_dir <- paste("output_", out_dir_suffix, sep = "")
dir.create(out_dir, showWarnings = FALSE)
```

===

Regional coordinate reference system taken from: 
http://spatialreference.org/ref/epsg/nad83-texas-state-mapping-system/proj4/

```{r, title = '{{ site.handouts[0] }}'}
CRS_reg <- "+proj=lcc +lat_1=27.41666666666667 +lat_2=34.91666666666666 +lat_0=31.16666666666667 +lon_0=-100 +x_0=1000000 +y_0=1000000 +ellps=GRS80 +datum=NAD83 +units=m +no_defs" 

file_format <- ".tif" 

NA_flag_val <- -9999

out_suffix <- "raster_classification" # output suffix for the files and output folder  

```

===

### Define input data file names

```{r, title = '{{ site.handouts[0] }}'}
infile_RITA_reflectance_date2 <- "mosaiced_MOD09A1_A2005273__006_reflectance_masked_RITA_reg_1km.tif"

infile_reg_outline <- "new_strata_rita_10282017" # Region outline and FEMA zones

infile_modis_bands_information <- "df_modis_band_info.txt" # MOD09 bands information.

nlcd_2006_filename <- "nlcd_2006_RITA.tif" # NLCD2006 Land cover data aggregated at ~ 1km.
```

===

## Part I: Read and map flooding from RITA hurricane 

===

### Display and explore the data

MOD09 raster image after hurricane Rita.

```{r, title = '{{ site.handouts[0] }}'}
r_after <- brick(file.path(in_dir, infile_RITA_reflectance_date2))
```

```{r}
head(r_after)
inMemory(r_after) # check if raster in memory
```

===

Read band information since it is more informative.

```{r, title = '{{ site.handouts[0] }}'}
df_modis_band_info <- read.table(file.path(in_dir, infile_modis_bands_information),
                                 sep=",", stringsAsFactors = F)
```

```{r}
df_modis_band_info
```

===

```{r, title = '{{ site.handouts[0] }}'}
names(r_after) <- df_modis_band_info$band_name
```

```{r}
names(r_after)
```

===

Define bands of interest: Use subset instead of $ if you want to wrap code into function
```{r, title = '{{ site.handouts[0] }}'}
raft_gr <- subset(r_after, "Green") 
raft_SWIR <- subset(r_after, "SWIR2")
raft_NIR <- subset(r_after, "NIR") 
raft_rd <- subset(r_after, "Red")
```

===

Manipulate bands of interest
```{r, title = '{{ site.handouts[0] }}'}
r_after_MNDWI <- (raft_gr - raft_SWIR) / (raft_gr + raft_SWIR)

plot(r_after_MNDWI, zlim = c(-1,1))
```

```{r, title = '{{ site.handouts[0] }}'}
r_after_NDVI <- (raft_NIR - raft_rd) / (raft_NIR + raft_rd)

plot(r_after_NDVI)
```

```{r}
inMemory(r_after_MNDWI) # check if raster in memory
inMemory(r_after_NDVI)
```

===

Write out rasters
```{r}
dataType(r_after_NDVI)
```

```{r, title = '{{ site.handouts[0] }}'}
data_type_str <- dataType(r_after_NDVI)

NAvalue(r_after_MNDWI) <- NA_flag_val # set NA value
out_filename_mn <- file.path(out_dir, paste0("mndwi_post_Rita","_", out_suffix, file_format))
writeRaster(r_after_MNDWI, filename = out_filename_mn, datatype = data_type_str,
            overwrite = T)
 
NAvalue(r_after_NDVI) <- NA_flag_val
out_filename_nd <- file.path(out_dir, paste0("ndvi_post_Rita","_", out_suffix, file_format))
writeRaster(r_after_NDVI, filename = out_filename_nd, datatype = data_type_str,
            overwrite = T)
```

===

Remove rasters from memory
```{r, title = '{{ site.handouts[0] }}'}
rm(r_after_MNDWI)
rm(r_after_NDVI)
```
Can view the .tif images written out in photo viewer.  

Read in rasters 
```{r, title = '{{ site.handouts[0] }}'}
r_after_MNDWI <- raster(out_filename_mn)
r_after_NDVI <- raster(out_filename_nd)
```

```{r}
inMemory(r_after_MNDWI) ; inMemory(r_after_NDVI)
```

===

### Groundtruthing data
Three levels of flooding (water content) determined via external grountruthing. 
Class 1: vegetation and other (small water fraction)
Class 2: Flooded vegetation
Class 3: Flooded area, or water (lake etc)

===

Read in a polygon for each class:

```{r, title = '{{ site.handouts[0] }}'}
class1_data_sf <- st_read(file.path(in_dir, "class1_sites")) 
class2_data_sf <- st_read(file.path(in_dir, "class2_sites"))
class3_data_sf <- st_read(file.path(in_dir, "class3_sites"))
```

```{r}
class1_data_sf
plot(r_after_MNDWI, zlim=c(-1,1)) 
plot(class3_data_sf["class_ID"], add = TRUE, color = NA, border = "blue", lwd = 3)
```

===

Combine polygons; note they should be in the same projection system.
```{r, title = '{{ site.handouts[0] }}'}
class_data_sf <- rbind(class1_data_sf, class2_data_sf, class3_data_sf)

class_data_sf$poly_ID <- 1:nrow(class_data_sf) # unique ID for each polygon
```

23 different polygons used as ground truth data.
```{r}
nrow(class_data_sf) 
```

===

Make new rasters and raster stack.
```{r, title = '{{ site.handouts[0] }}'}
r_x <- init(r_after, "x") # initializes a raster with coordinates x
r_y <- init(r_after, "y") # initializes a raster with coordiates y

r_stack <- stack(r_x, r_y, r_after, r_after_NDVI, r_after_MNDWI)
names(r_stack) <- c("x", "y", "Red", "NIR", "Blue", "Green", "SWIR1", "SWIR2", "SWIR3", "NDVI", "MNDWI")
```

```{r}
str(r_stack, max.level = 2)
```

```{r, title = '{{ site.handouts[0] }}'}
pixels_extracted_df <- extract(r_stack, class_data_sf, df=T)
```

```{r}
dim(pixels_extracted_df) 
```
We have 1547 pixels extracted.

===

Sometimes useful to store data as a csv: have to remove geometry column.
```{r, title = '{{ site.handouts[0] }}'}
class_data_df <- class_data_sf
st_geometry(class_data_df) <- NULL # This will coerce the sf object into a data.frame by removing the geometry
```

Make the dataframe used for classification.
```{r, title = '{{ site.handouts[0] }}'}
pixels_df <- merge(pixels_extracted_df, class_data_df, by.x="ID", by.y="poly_ID")
```

```{r}
head(pixels_df)
```

```{r, title = '{{ site.handouts[0] }}'}
table(pixels_df$class_ID) # count by class of pixels ground truth data
```

===

### Examining data used for the classification

Define geographic range.

```{r, title = '{{ site.handouts[0] }}'}
x_range <- range(pixels_df$Green, na.rm=T)
y_range <- range(pixels_df$NIR, na.rm=T)
```

Let's use a palette that reflects flooding or level of water.

```{r, title = '{{ site.handouts[0] }}'}
col_palette <- c("green", "lightblue", "darkblue")

```

===

Many ways of looking at these data:

```{r}
plot(NIR~Green, xlim=x_range, ylim=y_range, cex=0.3, col=col_palette[1], subset(pixels_df, class_ID==1))
points(NIR~Green, col=col_palette[2], cex=0.3, subset(pixels_df, class_ID==2))
points(NIR~Green, col=col_palette[3], cex=0.3, subset(pixels_df, class_ID==3))
names_vals <- c("vegetation","wetland","water")
legend("topleft", legend=names_vals,
       pt.cex=0.7, cex=0.7, col=col_palette,
       pch=20, # add circle symbol to line
       bty="n")
```

===

```{r, title = '{{ site.handouts[0] }}'}
plot(NDVI ~ MNDWI, xlim=c(-1,1), ylim=c(-1,1), cex=0.3, col=col_palette[1], subset(pixels_df, class_ID==1))
points(NDVI ~ MNDWI, cex=0.3, col=col_palette[2], subset(pixels_df, class_ID==2))
points(NDVI ~ MNDWI, cex=0.3, col=col_palette[3], subset(pixels_df, class_ID==3))
names_vals <- c("vegetation","wetland","water")
legend("topright", legend=names_vals,
       pt.cex=0.7,cex=0.7, col=col_palette,
       pch=20, # add circle symbol to line
       bty="n")
```

===

```{r, title = '{{ site.handouts[0] }}'}
rasterVis::histogram(r_after)
```

===

```{r, title = '{{ site.handouts[0] }}'}
pixels_df$class_ID <- factor(pixels_df$class_ID, levels = c(1,2,3), labels = names_vals)

boxplot(MNDWI~class_ID, pixels_df, xlab="category",
        main="Boxplot for MNDWI per class")
```

===

## Part II: Split data into training and testing datasets 

===

Let's keep 30% of data for testing for each class, and set a fixed seed for reproducibility.

```{r, title = '{{ site.handouts[0] }}'}
set.seed(100)
prop <- 0.3
table(pixels_df$class_ID)
```

===

```{r, title = '{{ site.handouts[0] }}'}
list_data_df <- vector("list", length=3)
level_labels <- names_vals

for(i in 1:3){
  data_df <- subset(pixels_df, class_ID==level_labels[i])
  data_df$pix_id <- 1:nrow(data_df)
  indices <- as.vector(createDataPartition(data_df$pix_id, p=(1-prop), list=F))
  data_df$training <- as.numeric(data_df$pix_id %in% indices)
  list_data_df[[i]] <- data_df
}

data_df <- do.call(rbind, list_data_df)
```

```{r}
dim(data_df)
head(data_df)
```

```{r, title = '{{ site.handouts[0] }}'}
data_training <- subset(data_df, training==1)
```

===

## Part III: Generate classification using classification tree (CART) and support vector machine (SVM)

===

### Classification and Regression Tree model (CART) 

Fit model using training data for CART.

```{r, title = '{{ site.handouts[0] }}'}
mod_rpart <- rpart(class_ID ~ Red + NIR + Blue + Green + SWIR1 + SWIR2 + SWIR3,
                   method="class", data=data_training)
```

Plot the fitted classification tree.

```{r, title = '{{ site.handouts[0] }}'}
plot(mod_rpart, uniform = TRUE, main = "Classification Tree", mar = c(0.1, 0.1, 0.1, 0.1))
text(mod_rpart, cex = .8)
```

===

Now predict the subset data based on the model.

```{r, title = '{{ site.handouts[0] }}', progress = FALSE, results = 'hide'}
raster_out_filename_rp <- file.path(out_dir, paste0("r_predicted_rpart_", out_suffix, file_format))

r_predicted_rpart <- predict(r_stack, mod_rpart, type='class', filename = raster_out_filename_rp,
                             progress = 'text', overwrite = T)
```

```{r}
plot(r_predicted_rpart)
```

```{r, title = '{{ site.handouts[0] }}'}
r_predicted_rpart <- ratify(r_predicted_rpart) # creates a raster attribute table (rat)
rat <- levels(r_predicted_rpart)[[1]]
rat$legend <- c("vegetation", "wetland", "water")
levels(r_predicted_rpart) <- rat

levelplot(r_predicted_rpart, maxpixels = 1e6,
          col.regions = c("green","blue","darkblue"),
          scales=list(draw=FALSE),
          main = "Classification Tree")
```

===

### Support Vector Machine classification (SVM)

Set class_ID as factor to generate classification.

```{r, title = '{{ site.handouts[0] }}'}
mod_svm <- svm(class_ID ~ Red + NIR + Blue + Green + SWIR1 + SWIR2 + SWIR3,
               data=data_training, method="C-classification",
               kernel="linear") # can be radial
```

```{r}
summary(mod_svm)
```

===

Now predict the subset data based on the model.

```{r, title = '{{ site.handouts[0] }}', progress = FALSE, results = 'hide'}
raster_out_filename_sv <- file.path(out_dir, paste0("r_predicted_svm_", out_suffix, file_format))
  
r_predicted_svm <- predict(r_stack, mod_svm, progress = 'text',
                           filename = raster_out_filename_sv, overwrite = T)
```

```{r, title = '{{ site.handouts[0] }}'}
plot(r_predicted_svm)
```

```{r, title = '{{ site.handouts[0] }}'}
rasterVis::histogram(r_predicted_svm)
```

```{r, title = '{{ site.handouts[0] }}'}
r_predicted_svm <- ratify(r_predicted_svm)
rat <- levels(r_predicted_svm)[[1]]
rat$legend <- c("vegetation", "wetland", "water")
levels(r_predicted_svm) <- rat

levelplot(r_predicted_svm, maxpixels = 1e6,
          col.regions = c("green","blue","darkblue"),
          scales = list(draw = FALSE),
          main = "SVM classification")
```

===

## Part IV: Assessment and comparison of model performance

Set up testing data:
Full dataset, let's use data points for testing.
```{r}
dim(data_df) 
```

```{r, title = '{{ site.handouts[0] }}'}
# omit values that contain NA, because may be problematic with SVM.
data_testing <- na.omit(subset(data_df, training==0)) 
```

```{r}
dim(data_testing)
```

===

Predict on testing data using rpart model fitted with testing data.

```{r, title = '{{ site.handouts[0] }}'}
testing_rpart <- predict(mod_rpart, data_testing, type='class')
```

Predict on testing data using SVM model fitted with testing data.

```{r, title = '{{ site.handouts[0] }}'}
testing_svm <- predict(mod_svm, data_testing, type='class')
```

Predicted number of pixels in each class:

```{r, title = '{{ site.handouts[0] }}'}
table(testing_rpart)
table(testing_svm)
```

===

### Generate confusion matrix to assess the accuracy of the image classification

More info on confusion matrices here: http://spatial-analyst.net/ILWIS/htm/ilwismen/confusion_matrix.htm

Groundtruth data: 

```{r}
table(data_testing$class_ID)
```

Classification model results:

```{r}
table(testing_rpart)
table(testing_svm)
```

To generate confusion matrix, need to use: 
testing_rpart: contains classification model prediction in the rows
data_testing$class_ID: this column contains groundtruth data

```{r, title = '{{ site.handouts[0] }}'}
tb_rpart <- table(testing_rpart, data_testing$class_ID)
tb_svm <- table(testing_svm, data_testing$class_ID)
```

Matrices identifying correct and incorrect classifications by models:

```{r}
tb_rpart
tb_svm
```

===

Accuracy (producer's accuracy): the fraction of correctly classified pixels compared to all pixels of that groundtruth class.

```{r, title = '{{ site.handouts[0] }}'}
tb_rpart[1]/sum(tb_rpart[,1]) # producer accuracy
tb_svm[1]/sum(tb_svm[,1]) # producer accuracy
```

Looking at accuracy more closely:

```{r, title = '{{ site.handouts[0] }}'}
# overall accuracy for rpart
sum(diag(tb_rpart))/sum(table(testing_rpart))

# overall accuracy for svm
sum(diag(tb_svm))/sum(table(testing_svm))
```

===

Generate more accurate measurements from functions in the caret package:

```{r, title = '{{ site.handouts[0] }}'}
accuracy_info_rpart <- confusionMatrix(testing_rpart, data_testing$class_ID, positive = NULL)
accuracy_info_svm <- confusionMatrix(testing_svm, data_testing$class_ID, positive = NULL)
```

Overall accuracy produced:

```{r}
accuracy_info_rpart$overall
```

```{r}
accuracy_info_svm$overall
```

Write out the results:

```{r, title = '{{ site.handouts[0] }}'}
table_out_filename <- file.path(out_dir, "confusion_matrix_rpart.txt")
write.table(accuracy_info_rpart$table, table_out_filename, sep=",")

table_out_filename <- file.path(out_dir, "confusion_matrix_svm.txt")
write.table(accuracy_info_svm$table, table_out_filename, sep=",")
```
