---
editor_options: 
  chunk_output_type: console
excerpt: Raster Classification using flood mapping analysis
---

```{r, include=FALSE, echo=FALSE, results='hide'}
library(raster) # raster functionalities
library(sf) # spatial objects classes
library(rpart) # regression and classification trees 
library(e1071) # support vector machine
library(rasterVis) # raster visualization operations
library(caret) # complex regression and classification models
```


### Set up arguments and parameters

```{r, title = '{{ site.handouts[0] }}'}
in_dir <- "data"

out_dir_suffix <- "rsc_lesson"
out_dir <- paste("output_", out_dir_suffix, sep="")
dir.create(out_dir, showWarnings = FALSE)
```

===

Regional coordinate reference system taken from: 
http://spatialreference.org/ref/epsg/nad83-texas-state-mapping-system/proj4/

```{r, title = '{{ site.handouts[0] }}'}
CRS_reg <- "+proj=lcc +lat_1=27.41666666666667 +lat_2=34.91666666666666 +lat_0=31.16666666666667 +lon_0=-100 +x_0=1000000 +y_0=1000000 +ellps=GRS80 +datum=NAD83 +units=m +no_defs" 

file_format <- ".tif" 

NA_flag_val <- -9999

out_suffix <- "raster_classification" # output suffix for the files and output folder  

```

===

### Define input data file names

```{r, title = '{{ site.handouts[0] }}'}
infile_RITA_reflectance_date2 <- "mosaiced_MOD09A1_A2005273__006_reflectance_masked_RITA_reg_1km.tif"

infile_reg_outline <- "new_strata_rita_10282017" # Region outline and FEMA zones

infile_modis_bands_information <- "df_modis_band_info.txt" # MOD09 bands information.

nlcd_2006_filename <- "nlcd_2006_RITA.tif" # NLCD2006 Land cover data aggregated at ~ 1km.
```

===

## Part I: Read and map flooding from RITA hurricane 

### Display and explore the data

MOD09 raster image after hurricane Rita.
```{r, title = '{{ site.handouts[0] }}'}
r_after <- brick(file.path(in_dir,infile_RITA_reflectance_date2))
```

===

Read band information since it is more informative.

```{r, title = '{{ site.handouts[0] }}'}
df_modis_band_info <- read.table(file.path(in_dir, infile_modis_bands_information),
                                 sep=",",
                                 stringsAsFactors = F)

print(df_modis_band_info)

df_modis_band_info$band_number <- c(3,4,1,2,5,6,7)

band_refl_order <- df_modis_band_info$band_number

names(r_after) <- df_modis_band_info$band_name
```

===

Note: Use subset instead of $ if you want to wrap code into function.
```{r, title = '{{ site.handouts[0] }}'}
r_after_MNDWI <- (subset(r_after,"Green") - subset(r_after,"SWIR2")) / (subset(r_after,"Green") + subset(r_after,"SWIR2"))
plot(r_after_MNDWI,zlim=c(-1,1))

r_after_NDVI <- (subset(r_after,"NIR") - subset(r_after,"Red")) / (subset(r_after,"NIR") + subset(r_after,"Red"))
plot(r_after_NDVI)

data_type_str <- dataType(r_after_NDVI)


NAvalue(r_after_MNDWI) <- NA_flag_val
out_filename <- paste("mndwi_post_Rita","_", out_suffix, file_format, sep="")
out_filename <- file.path(out_dir, out_filename)
writeRaster(r_after_MNDWI,
            filename=out_filename,
            datatype=data_type_str,
            overwrite=T)
 
NAvalue(r_after_NDVI) <- NA_flag_val
out_filename <- paste("ndvi_post_Rita","_", out_suffix, file_format, sep="")
out_filename <- file.path(out_dir, out_filename)
writeRaster(r_after_NDVI,
            filename=out_filename,
            datatype=data_type_str,
            overwrite=T)
```

===

Class 1) vegetation and other (small water fraction)
Class 2) Flooded vegetation
Class 3) Flooded area, or water (lake etc)

Make a dataframe for each class:
```{r, title = '{{ site.handouts[0] }}'}
class1_data_sf <- st_read(file.path(in_dir, "class1_sites"))
class2_data_sf <- st_read(file.path(in_dir, "class2_sites"))
class3_data_sf <- st_read(file.path(in_dir, "class3_sites"))
```

===

Combine objects; note they should be in the same projection system.
```{r, title = '{{ site.handouts[0] }}'}
class_data_sf <- rbind(class1_data_sf, class2_data_sf, class3_data_sf)

class_data_sf$poly_ID <- 1:nrow(class_data_sf) # unique ID for each polygon
```

23 different polygons used as ground truth data.
```{r}
nrow(class_data_sf) 
```

===

```{r, title = '{{ site.handouts[0] }}'}
class_data_sp <- as(class_data_sf, "Spatial")

r_x <- init(r_after,"x") # raster with coordinates x
r_y <- init(r_after,"x") # raster with coordiates y

r_stack <- stack(r_x, r_y, r_after, r_after_NDVI, r_after_MNDWI)
names(r_stack) <- c("x","y","Red","NIR","Blue","Green","SWIR1","SWIR2","SWIR3","NDVI","MNDWI")
pixels_extracted_df <- extract(r_stack, class_data_sf, df=T)
```

```{r}
dim(pixels_extracted_df) 
```
We have 1547 pixels extracted.

===

```{r, title = '{{ site.handouts[0] }}'}
class_data_df <- class_data_sf
st_geometry(class_data_df) <- NULL # This will coerce the sf object into a data.frame

pixels_df <- merge(pixels_extracted_df, class_data_df, by.x="ID", by.y="poly_ID")
```

```{r}
head(pixels_df)
```

```{r, title = '{{ site.handouts[0] }}'}
table(pixels_df$class_ID) # count by class of pixels ground truth data
```

===

### Examining site data used for the classification
Water
```{r, title = '{{ site.handouts[0] }}'}
x_range <- range(pixels_df$Green,na.rm=T)
y_range <- range(pixels_df$NIR,na.rm=T)

plot(NIR~Green,xlim=x_range,ylim=y_range,cex=0.2,col="blue",subset(pixels_df,class_ID==1))
points(NIR~Green,col="green",cex=0.2,subset(pixels_df,class_ID==2))
points(NIR~Green,col="red",cex=0.2,subset(pixels_df,class_ID==3))
names_vals <- c("water class 1","water class 2","water class 3")
legend("topleft",legend=names_vals,
       pt.cex=0.7,cex=0.7,col=c("blue","green","red"),
       pch=20, # add circle symbol to line
       bty="n")
```

===

Let's use a palette that reflects wetness or level of water. 
```{r, title = '{{ site.handouts[0] }}'}
col_palette <- c("green","blue","darkblue")

plot(NDVI ~ MNDWI,
     xlim=c(-1,1),ylim=c(-1,1),
     cex=0.2,
     col=col_palette[1],
     subset(pixels_df,class_ID==1))
points(NDVI ~ MNDWI,
       cex=0.2,
       col=col_palette[2],
       subset(pixels_df,class_ID==2))
points(NDVI ~ MNDWI,
       cex=0.2,
       col=col_palette[3],
       subset(pixels_df,class_ID==3))
names_vals <- c("vegetation","wetland","water")
legend("topright",legend=names_vals,
       pt.cex=0.7,cex=0.7,col=col_palette,
       pch=20, # add circle symbol to line
       bty="n")

rasterVis::histogram(r_after)

pixels_df$class_ID <- factor(pixels_df$class_ID,
                      levels = c(1,2,3),
                      labels = names_vals)

boxplot(MNDWI~class_ID,
        pixels_df,
        xlab="category",
        main="Boxplot for MNDWI per class")
```

===

## Part II: Split data into training and testing datasets 

Let's keep 30% of data for testing for each class.
```{r, title = '{{ site.handouts[0] }}'}
pixels_df$pix_ID <- 1:nrow(pixels_df)
prop <- 0.3
table(pixels_df$class_ID)
set.seed(100) # set random seed for reproducibility
```

===

This is for one class: better to do this as a function but we use a loop for clarity here.
```{r, title = '{{ site.handouts[0] }}'}
list_data_df <- vector("list", length=3)
level_labels <- names_vals

for(i in 1:3){
  data_df <- subset(pixels_df, class_ID==level_labels[i])
  data_df$pix_id <- 1:nrow(data_df)
  indices <- as.vector(createDataPartition(data_df$pix_ID, p=0.7, list=F))
  data_df$training <- as.numeric(data_df$pix_id %in% indices)
  list_data_df[[i]] <- data_df
}

data_df <- do.call(rbind, list_data_df)
```

```{r}
dim(data_df)
head(data_df)
```

```{r, title = '{{ site.handouts[0] }}'}
data_training <- subset(data_df, training==1)
```

===

## Part III: Generate classification using classification tree (CART) and support vector machine (SVM)

### Classification and Regression Tree model (CART) 

Fit model using training data for CART.
```{r, title = '{{ site.handouts[0] }}'}
mod_rpart <- rpart(class_ID ~ Red + NIR + Blue + Green + SWIR1 + SWIR2 + SWIR3,
                   method="class", 
                   data=data_training)
```

Plot the fitted classification tree.
```{r, title = '{{ site.handouts[0] }}'}
plot(mod_rpart, uniform = TRUE, main = "Classification Tree", mar = c(0.1, 0.1, 0.1, 0.1))
text(mod_rpart, cex = .8)
```

===

Now predict the subset data based on the model; prediction for entire area takes more time.
```{r, title = '{{ site.handouts[0] }}'}
raster_out_filename <- paste0("r_predicted_rpart_", out_suffix, file_format)
raster_out_filename <- file.path(out_dir, raster_out_filename)

r_predicted_rpart <- predict(r_stack, mod_rpart, 
                             type='class',
                             filename=raster_out_filename,
                             progress = 'text',
                             overwrite=T)

plot(r_predicted_rpart)

r_predicted_rpart <- ratify(r_predicted_rpart)
rat <- levels(r_predicted_rpart)[[1]]
rat$legend <- c("vegetation","wetland","water")
levels(r_predicted_rpart) <- rat

levelplot(r_predicted_rpart, maxpixels = 1e6,
          col.regions = c("green","blue","darkblue"),
          scales=list(draw=FALSE),
          main = "Classification Tree")
```

===

### Support Vector Machine classification (SVM)

Set class_ID as factor to generate classification
```{r, title = '{{ site.handouts[0] }}'}
mod_svm <- svm(class_ID ~ Red +NIR + Blue + Green + SWIR1 + SWIR2 + SWIR3,
               data=data_training,
               method="C-classification",
               kernel="linear") # can be radial

summary(mod_svm)
```

===

Now predict the subset data based on the model; prediction for entire area takes longer time
```{r, title = '{{ site.handouts[0] }}'}
raster_out_filename <- paste0("r_predicted_svm_", out_suffix, file_format)
raster_out_filename <- file.path(out_dir, raster_out_filename)
  
r_predicted_svm <- predict(r_stack, mod_svm,
                           progress = 'text',
                           filename = raster_out_filename,
                           overwrite = T)

plot(r_predicted_svm)

rasterVis::histogram(r_predicted_svm)

r_predicted_svm <- ratify(r_predicted_svm)
rat <- levels(r_predicted_svm)[[1]]
rat$legend <- c("vegetation", "wetland", "water")
levels(r_predicted_svm) <- rat

levelplot(r_predicted_svm, maxpixels = 1e6,
          col.regions = c("green","blue","darkblue"),
          scales = list(draw = FALSE),
          main = "SVM classification")
```

===

## Part IV: Assessment and comparison of model performance

Set up testing data:
```{r, title = '{{ site.handouts[0] }}'}
dim(data_df) # full dataset, let's use data points for testing

# omit values that contain NA, because may be problematic with SVM.
data_testing <- na.omit(subset(data_df, training==0)) 
```

```{r}
dim(data_testing)
```

===

Predict on testing data using rpart model fitted with training data.
```{r, title = '{{ site.handouts[0] }}'}
testing_rpart <- predict(mod_rpart, data_testing, type='class')
```

Predict on testing data using SVM model fitted with training data
```{r, title = '{{ site.handouts[0] }}'}
testing_svm <- predict(mod_svm, data_testing, type='class')
```

Predicted classes:
```{r, title = '{{ site.handouts[0] }}'}
table(testing_svm)
```

===

### Generate confusion matrix to assess the accuracy of the image classification
More info on confusion matrices here: http://spatial-analyst.net/ILWIS/htm/ilwismen/confusion_matrix.htm

Need to use: 
testing_rpart: contains map prediction in the rows
data_testing$class_ID: this column contains ground truth data 
```{r, title = '{{ site.handouts[0] }}'}
tb_rpart <- table(testing_rpart, data_testing$class_ID)
tb_svm <- table(testing_svm, data_testing$class_ID)

table(testing_rpart) # classification, map results
table(data_testing$class_ID) # reference, ground truth in columns
```

===

Accuracy (producer's accuracy): the fraction of correctly classified pixels compared to all pixels 
of that ground truth class. 
```{r, title = '{{ site.handouts[0] }}'}
tb_rpart[1]/sum(tb_rpart[,1]) # producer accuracy
tb_svm[1]/sum(tb_svm[,1]) # producer accuracy
```

Looking at accuracy more closely:
```{r, title = '{{ site.handouts[0] }}'}
# overall accuracy for svm
sum(diag(tb_svm))/sum(table(testing_svm))

# overall accuracy for rpart
sum(diag(tb_rpart))/sum(table(testing_rpart))
```

===

Generate more accurate measurements from functions in the caret package:

```{r, title = '{{ site.handouts[0] }}'}
accuracy_info_svm <- confusionMatrix(testing_svm, data_testing$class_ID, positive = NULL)
accuracy_info_rpart <- confusionMatrix(testing_rpart, data_testing$class_ID, positive = NULL)

accuracy_info_rpart$overall  # overall accuracy produced
accuracy_info_svm$overall
```

Write out the results:

```{r, title = '{{ site.handouts[0] }}'}
table_out_filename <- file.path(out_dir, "confusion_matrix_rpart.txt")
write.table(accuracy_info_rpart$table, table_out_filename, sep=",")

table_out_filename <- file.path(out_dir, "confusion_matrix_svm.txt")
write.table(accuracy_info_svm$table, table_out_filename, sep=",")
```
